 

### 2021-07-05 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

**Rising temperatures are likely to have differential impacts on trees of different size and canopy position, in large part because small understory trees exist in a microenvironment that is substantially buffered by the more exposed canopy trees [@zellweger_seasonal_2019; @davisMicroclimaticBufferingForests2019].**
Forests are vertically and horizontally stratified, with trees of different sizes through various successional stages. 
Overstory trees form canopies that play a crucial role in moderating forest macroclimates [@ozanneBiodiversityMeetsAtmosphere2003; @nakamuraForestsTheirCanopies2017], as well as microclimates by buffering understories from extreme temperatures and other macroclimatic conditions [@zellweger_seasonal_2019].
This creates a vertical stratification of biophysical environmental conditions such as temperature, light, wind, humidity and CO~2~ that influences leaf traits, thermoregulation and metabolism along the gradient, with implications for whole plant performance [@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016].
Despite the fact that this vertical gradient inevitably shapes nearly every aspect of plant metabolism, demography, and ecology, we lack comprehensive understanding of these gradients [but see @niinemetsWithincanopyVariationRate2004a].
Importantly, this limits our ability to understand how warming temperatures will affect leaf-level metabolism, whole-plant performance, and, in turn, forest ecology and biodiversity, energy balance, ecosystem function, and biosphere-atmosphere interactions.

@@

**Air temperature, $T_{air}$, often shows little variation across the vertical gradient, but under certain circumstances it can be significantly buffered by forest canopies (Figs. `r fig_NEON_vertical`, `r SIfig_NEON_all`).**
The $T_{air}$ gradient along canopies and understories varies with factors such as radiation fluxes, local topography, vertical air mixing, cloud cover and vegetation [*Defrenne et al. 2021*]. 
Typically, dense canopies buffer understories from maximum $T_{air}$ relatively more than open canopies -- *i.e.*, dense canopy understories can have cooler maximum daytime $T_{air}$ and warmer minimum $T_{air}$ during the night than open canopy understories or nearby clearings [*Defrenne et al. 2021*; @vonarxSpatiotemporalEffectsForest2012a; @defrenneGlobalBufferingTemperatures2019; @davis_microclimatic_2019; @zellweger_seasonal_2019; @ramboCanopyMicroclimateResponse2009]. 
Notably, this pattern can be reversed in open canopy forests or sparse stands, where below-canopy maximum $T_{air}$ can be warmer than canopy $T_{air}$ due to turbulent air mixing and thermal radiative flux from the soil and the canopy that is intercepted by lower-canopy layers, analogous to 'canopy greenhouse effect' [e.g., OSBS, Figs. `r fig_NEON_vertical`g, `r SIfig_NEON_all`b, @banerjeeExplainingConvectorEffect2017; @curtis_intracanopy_2019;@zweifelMiddayStomatalClosure2002; @hadleyInfluenceKrummholzMat1987; @ramboCanopyMicroclimateResponse2009; @hardwickRelationshipLeafArea2015].

>>>>

**Rising temperatures are likely to have differential impacts on trees of different size and canopy position, in large part because small understory trees exist in a microenvironment that is substantially buffered by the more exposed canopy trees [@zellweger_seasonal_2019; @davisMicroclimaticBufferingForests2019].**
Forests are vertically and horizontally stratified, with trees of different sizes through various successional stages. 
Overstory trees form canopies that play a crucial role in moderating forest macroclimates [@ozanneBiodiversityMeetsAtmosphere2003; @nakamuraForestsTheirCanopies2017], as well as microclimates by buffering understories from extreme temperatures and other macroclimatic conditions [@zellweger_seasonal_2019].
This creates a vertical stratification of biophysical environmental conditions such as temperature, light, wind, humidity and CO~2~ that influences leaf traits, thermoregulation and metabolism along the gradient, with implications for whole plant performance [@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016].
Despite the fact that this vertical gradient inevitably shapes nearly every aspect of plant metabolism, demography, and ecology, we lack comprehensive understanding of these gradients [but see @niinemetsPhotosyntheticAcclimationSimultaneous2004].
Importantly, this limits our ability to understand how warming temperatures will affect leaf-level metabolism, whole-plant performance, and, in turn, forest ecology and biodiversity, energy balance, ecosystem function, and biosphere-atmosphere interactions.

@@

**Air temperature, $T_{air}$, often shows little variation across the vertical gradient, but under certain circumstances it can be significantly buffered by forest canopies (Figs. `r fig_NEON_vertical`, `r SIfig_NEON_all`).**
The $T_{air}$ gradient along canopies and understories varies with factors such as radiation fluxes, local topography, vertical air mixing, cloud cover and vegetation [*Defrenne et al. 2021*]. 
Typically, dense canopies buffer understories from maximum $T_{air}$ relatively more than open canopies -- *i.e.*, dense canopy understories can have cooler maximum daytime $T_{air}$ and warmer minimum $T_{air}$ during the night than open canopy understories or nearby clearings [*Defrenne et al. 2021*; @vonarxSpatiotemporalEffectsForest2012a; @defrenneGlobalBufferingTemperatures2019; @davisMicroclimaticBufferingForests2019; @zellweger_seasonal_2019; @ramboCanopyMicroclimateResponse2009]. 
Notably, this pattern can be reversed in open canopy forests or sparse stands, where below-canopy maximum $T_{air}$ can be warmer than canopy $T_{air}$ due to turbulent air mixing and thermal radiative flux from the soil and the canopy that is intercepted by lower-canopy layers, analogous to 'canopy greenhouse effect' [e.g., OSBS, Figs. `r fig_NEON_vertical`g, `r SIfig_NEON_all`b, @banerjeeExplainingConvectorEffect2017; @curtis_intracanopy_2019;@zweifelMiddayStomatalClosure2002; @hadleyInfluenceKrummholzMat1987; @ramboCanopyMicroclimateResponse2009; @hardwickRelationshipLeafArea2015].
